---
title: "Phylogeneitc structure variables"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This script uses some functions from the an R package `toolkitSeqTree` to generate variables that . Such variables can be used in statistical analyses to account for the non-independence of taxa that are the result of an evolutionary process and associated issues with repeated measurements.

```{r}
library(toolkitSeqTree)
library(ggtree)
library(ggplot2)
library(dplyr)
```


```{r}
tree <- treeio::read.beast('extdata/h1n1_ref_viruses.trees')
tree_dat <- fortify(tree)
ggtree(tree)
head(tree_dat)
```

The relationships between nodes of the phylogeny are contained in `tree_dat` columns `tree_dat$node` and `tree_dat$parent`. For a bifurcating tree, each internal node of the phylogeny is the parent of two nodes.

The exception to this rule is that the root of the phylogeny is the parent of three nodes including itself, so for the root of the phylogeny, `tree_dat$node == tree_dat$parent`:

```{r}
# plot tree with root highlighted
ggtree(tree) +
  geom_point(data = subset(tree_dat, node == parent),
             col = 'dodgerblue', size = 4)
```

### MRCA of two taxa or nodes.

The function `nodes_relationship()` will take two taxa labels (or alternatively two node numbers - which can include internal nodes) and with the argument `relationship = 'mrca'` will identify the node representing their most recent common ancestor (MRCA).

```{r}
# Pair of taxa of interest are selected and function 
taxa_interested <- c('H1N1_A_WUHAN_371_1995',
                     'H1N1_A_FUKUSHIMA_141_2006')
taxa_mrca <- nodes_relationship(tree_dat, taxa = taxa_interested,
                                relationship = 'mrca')

# Highlight taxa_interested as blue circles and MRCA as red diamond 
ggtree(tree) +
  geom_point(data = subset(tree_dat, label %in% taxa_interested),
             col = 'dodgerblue', size = 4) +
  geom_point(data = subset(tree_dat, node %in% taxa_mrca),
             col = 'firebrick2', shape = 18, size = 4)
```

### Path between nodes

It may be useful for some analyses to define a path between two nodes (probably between two tips in most cases). The nodes  that form this path lie between the two nodes and their MRCA.

This set of nodes (it may be more intuitive to consider the branches) forming this path is identified from the ancestors of the two nodes/taxa identified using `node_ancestors()`. The set forming the path is the set of nodes/branches which are in either of the sets of ancestors, but not their intersection (known as the symmetric difference or disjunctive union of the two sets). 

It will usually make sense to complete the path by supplementing this set of nodes with the node/branch connecting the target nodes using the `node_ancestors()` argument `keep_target_node = TRUE`.

```{r}
# ID ancestors of two randomly chosen taxa 
taxa_interested <- c('H1N1_A_WUHAN_371_1995',
                     'H1N1_A_FUKUSHIMA_141_2006')
ancestors_1 <- node_ancestors(tree_dat, taxa_interested[1],
                              keep_target_node = TRUE)
ancestors_2 <- node_ancestors(tree_dat, taxa_interested[2],
                              keep_target_node = TRUE)

# Nodes in the path between two taxa are identified as the difference
# between the union and the intersect of the two sets of ancestors
path_nodes <- setdiff(union(ancestors_1, ancestors_2),
                      intersect(ancestors_1, ancestors_2))

# Highlight path branches as thicker and red
ggtree(tree, aes(color = node %in% path_nodes,
                 size = node %in% path_nodes)) +
  scale_color_manual(values = c('black', 'firebrick2'), guide = 'none') +
  scale_size_manual(values = c(0.5, 1), guide = 'none') +
  geom_point(data = subset(tree_dat, label %in% taxa_interested),
             col = 'dodgerblue', size = 4)
```

Alternatively, the path between two taxa can be identified using `nodes_relationship()` with the relationship argument set to path (`relationship = 'path'`), which is also the default option.

```{r}
path_nodes

nodes_relationship(tree_dat, taxa = taxa_interested,
                   relationship = 'path')
```

## Creating variables

In an analysis exploring variation in some trait associated with a property of each taxa or perhaps measurements that reflect the similarity of taxa. To explore variation in a trait measured per taxa, it may be sensible to use variables representing whether each node/branch exists in the evolutionary history of that taxa (i.e. is the node/branch identified as an ancestor using `node_ancestors()`, or not). To explore variation in a trait measured between pairs of taxa, it may be sensible to to use variables representing whether each node/branch separates the pair in a path drawn through the tree along branches of the phylogeny. Variables of this nature can be classified using `nodes_relationship()` with `relationship = 'path'`.

### Difference variables

Logical variable for each node/branch and whether or not it is an ancestor of each taxa. This generates a set of variables with one observation per combination of taxa.

To get an idea of how this variable works, first select a viewpoint virus then look for a variable for a node/branch, for which other taxa does this node/branch fall on the path between them (i.e. diff variable == TRUE)

```{r}
# regenerate clean version of tree_dat
tree_dat <- fortify(tree)

# merge output from generate_branch_vars() with tree_dat
dat <- generate_branch_vars(tree_dat, variable_type = 'difference')

# subset this df do that only one virus is present as 'label_1' then
# left_join() using 'label_2'
taxa_subset <- subset(dat, label_1 == 'H1N1_A_FUKUSHIMA_141_2006')
tree_dat <- left_join(tree_dat, taxa_subset, by = c('label' = 'label_2'))

ggtree(tree) +
  # Viewpoint virus highlighted in red tip node
  geom_point(data = subset(tree_dat,
                           label == 'H1N1_A_FUKUSHIMA_141_2006'),
             col = 'firebrick3', size = 4) +
  # branch/node assoc. variable diff_48 highlighted with blue diamond
  geom_point(data = subset(tree_dat, node == 48),
             col = 'dodgerblue', shape = 18, size = 4) +
  # colour/shape of tip nodes depend on whether the combination of them and 
  # the viewpoint virus are separated by the branch
  geom_point(data = subset(tree_dat, isTip == TRUE),
             aes(col = diff_48, shape = diff_48), size = 2) +
  scale_color_manual(values = c('forestgreen', 'goldenrod'))
```

The above plot shows a small clade of viruses that are separated from the taxa labelled `'H1N1_A_FUKUSHIMA_141_2006'` for which `diff_48 == TRUE`. These branches are separated from all other viruses in the tree by branch/node 48. If a virus within this clade of 4 taxa is used as the viewpoint, the opposite will be true and `diff_48 == TRUE`
